# Principles of using `TMB` for MLE

## Setting up `TMB`
In order to run the scripts and example code, you first need to set up TMB by going through the following steps:

- Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
- Install the R-package `TMB` [@R-TMB] by executing the following code in R:

```{r 2install_TMB, eval=FALSE}
install.packages("TMB")
```

- (Optional) Setup error debugging in `RStudio` by running the command  `TMB:::setupRstudio()`


Advanced use is detailed in https://kaskr.github.io/adcomp/_book/Tutorial.html

## Getting started

Let $\b{x}$ and $\b{y}$ denote the predictor and response vector, respectively, both of length $n$.
For a simple linear regression model with intercept $a$ and slope $b$, the negative log-likelihood equals
\begin{equation*}
- \log L(a, b, \sigma^2) = - \sum_{i=1}^n \log(\phi(y_i; a + bx_i, \sigma^2)),
\end{equation*}
where $\phi(\cdot; \mu, \sigma^2)$ corresponds to the density function of the univariate normal distribution with mean $\mu$ and variance $\sigma^2$.

The use of `TMB` requires the (negative) log-likelihood function to be coded in C++ under a specific template, which is then loaded into R.
The minimization of this function and other post-processing procedures are all carried out in `R`.
Therefore, we require two files.\
The first file, named *[code/linreg.cpp](code/linreg.cpp)*, is written in C++ and defines the objective function, i.e. the negative log-likelihood (nll) function of the linear model.

```{Rcpp 2linreg.cpp, code=readLines("code/linreg.cpp"), eval=FALSE}
```

The second file needed is written in `R` and serves for compiling the nll function defined above and carrying out the estimation procedure by numerical optimization of the nll function.
The .R file (shown below) carries out the compilation of the C++ file and minimization of the nll function:
```{r 2init-linreg}
# Loading TMB package
library(TMB)
# Compilation. The compiler returns 0 if the compilation of
# the cpp file was successful
TMB::compile("code/linreg.cpp")
# Dynamic loading of the compiled cpp file
dyn.load(dynlib("code/linreg"))
# Generate the data for our test sample
set.seed(123)
data <- list(y = rnorm(20) + 1:20, x = 1:20)
parameters <- list(a = 0, b = 0, tsigma = 0)
# Instruct TMB to create the likelihood function
obj_linreg <- MakeADFun(data, parameters, DLL = "linreg",
                        silent = TRUE)
# Optimization of the objective function with nlminb
mod_linreg <- nlminb(obj_linreg$par, obj_linreg$fn,
                     obj_linreg$gr,
                     obj_linreg$he)
mod_linreg$par
```

Now that the optimization is taken care of, we can display the estimates with standard errors using the `sdreport` function.
```{r sdreport}
sdreport(obj_linreg)
```
If we use `summary` on this object, we also get the variables we have passed to `ADREPORT` in the *[code/linreg.cpp](code/linreg.cpp)* file.
In this example, this is only the residual standard deviation;  `sigma`.
```{r 2summary_sdreport}
summary(sdreport(obj_linreg))
```
The `select` argument restricts the output to variables passed
by `ADREPORT(variable_name);` in the cpp file.
As we will see in [Wald-type], this lets us derive confidence intervals for these natural parameters easily.
```{r 2summary_sdreport_report}
summary(sdreport(obj_linreg), select = "report")
```

Certainly, you would not build a TMB model to fit a linear regression. We can use standard R functions for that. Therefore, we use `stats::lm` on the same data and compare the estimates to those obtained by `TMB`.
```{r 2lm}
rbind(
  "lm"  = lm(y ~ x, data = data)$coef, # linear regression using R
  "TMB" = mod_linreg$par[1:2] # intercept and slope from TMB fit
)
```
As we can see, the parameter estimates are exactly the same.

## Extending the `C++` nll
### Minimal example
Sometimes it is useful to write subroutines as a separate function that can be used within your TMB likelihood function.
This can increase readability of your code and reduce the number of lines of code in your main cpp file.
Writing extra files to define functions compatible with `TMB` requires some care, as these must follow `TMB`'s template.
<!-- It is noteworthy that `linreg.cpp` doesn't have to contain many lines of code, as they can be distributed in other files and loaded. -->

To illustrate how this works, we add a separate function to the *[linreg.cpp](code/linreg.cpp)* example.
The following function does not do anything meaningful, but is just meant to show you how you can add write an additional function and load it into your C++.
We start by writing the subroutine function as a separate file called *[functions/utils_linreg.cpp](functions/utils_linreg.cpp)*.
```{Rcpp 2utils_linreg_extended.cpp, code=readLines("functions/utils_linreg_extended.cpp"), eval=FALSE}
```
We then import it into [*code/linreg_extended.cpp*](code/linreg_extended.cpp). 
```{Rcpp 2utils_linreg_extended.cpp, code=readLines("code/linreg_extended.cpp"), eval=FALSE}
```
And eventually we can use it in R, as shown in this minimal example.
```{r 2linreg_extended}
# Loading TMB package
library(TMB)
# Compilation. The compiler returns 0 if the compilation of
# the cpp file was successful
TMB::compile("code/linreg_extended.cpp")
# Dynamic loading of the compiled cpp file
dyn.load(dynlib("code/linreg_extended"))
# Generate the data for our test sample
set.seed(123)
sigma <- 0.6
data <- list(y = rnorm(20, sd = sigma) + 1:20, x = 1:20)
tsigma <- log(sigma / (1 - sigma)) # Logit transform
parameters <- list(a = 0, b = 0, tsigma = 0.1)
# Instruct TMB to create the likelihood function
obj_linreg <- MakeADFun(data, parameters, DLL = "linreg_extended",
                        silent = TRUE)
# Optimization of the objective function with nlminb
mod_linreg <- nlminb(obj_linreg$par, obj_linreg$fn,
                     obj_linreg$gr,
                     obj_linreg$he)
# Objects returned by ADREPORT() in C++
summary(sdreport(obj_linreg), select = "report")
# Object
obj_linreg$report()
```

Note that when you are writing the `C++` nll file, compiling the file again may lead to an error.
This is because the dynamic library (the `.dll` file) is already loaded, and cannot be overwritten.
To prevent this, it is useful to either to manually unload the file manually via the code
```{r dyn_unload, eval=FALSE}
dyn.unload(dynlib("code/linreg_extended"))
```
or to restart the `R` session via the menu `Session->Restart R` (shortcut `Ctrl+Shift+F10` on Windows).
Note that restarting the session unload all packages from the session, and will require you to load the `TMB` package again.

