<!-- ```{r 3import-files, echo = FALSE, cache = FALSE} -->
<!-- library(knitr) -->
<!-- setwd(dir = "../") -->
<!-- # suppressMessages(source("code/main.R")) -->
<!-- knitr::read_chunk('functions/utils.R') -->
<!-- ``` -->

# Parameter estimation techniques

## Utility functions

### In `TMB`

Defining the negative log-likelihood function requires transforming the working parameters into their natural form.
We define the function `gamma_w2n` to perform this transformation.

`stat_dist` handles computing the stationary distribution, although `delta_w2n` can be used if a non-stationary distribution is provided.

They are defined

```{Rcpp 3utils.cpp, code=readLines("functions/utils.cpp"), eval=FALSE}
```

Transforming the Poisson means into their natural form can be done simply with the `exp` function and doesn't require a dedicated function.

### In `R`
#### Parameters
While `TMB` requires functions to transform working parameters to their natural form, pre-processing in `R` requires the opposite.

Functions to achieve this are available in [@zucchini, p. ~51] and are displayed here for convenience.

```{r pois_HMM_pn2pw}
```

This can be broken down into sub-functions if necessary. For the $\gamma$ part, we introduce `gamma_n2w` below.
```{r gamma_n2w}
```

In the case where a non-stationary distribution is specified, transforming $\b{\delta}$ is necessary. For this we use the `delta_n2w` function.
```{r delta_n2w}
```

When assuming a stationary distribution, computing it can be done via the following `stat_dist` function.
```{r stat_dist}
```

#### Label switching
As the model gets estimated each time, we do not impose by default an order for the states. This can lead to the label switching problem, where states aren't ordered the same way in each model.
The issue can be relevant when comparing results of different optimizers, initial parameters, or classes of models. 

To address the problem, we reorder the states by ascending Poisson means.
Sorting the means is direct, however re-ordering the TPM is not as straightforward.
To do so, we take the permutations of the states given by the sorted Poisson means, and permute each row index and column index to its new value.
A function to achieve this is named `pois_HMM_label_order` and presented below. 
```{r pois_HMM_label_order}
```

We will go through an example to better understand the process.
For readability, the TPM is filled with standard row and column indexes instead of probabilities.
```{r 3relabel}
lambda <- c(3, 1, 2)
gamma <- matrix(c(11, 12, 13,
                  21, 22, 23,
                  31, 32, 33), byrow = TRUE, ncol = 3)
pois_HMM_label_order(m = 3, lambda, gamma)
```

State 1 has been relabeled state 3, state 3 became state 2, and state 2 became state 1.

Another way to address this issue is by parametrizing in terms of non-negative increments $\lambda_j - \lambda_{j-1}$ with $\lambda_0 \equiv 0$, as explained by [@zucchini Section 7.1.1 p. 112].
However, [@bulla, Section 3.2 p. 7] shows this can impose optimization issues. They write "We find that, over all series, the simplest parameterization, i.e., the use of log-transformed state-dependent parameters, leads to the best results as regards the number of failures and the convergence to the global maximum".

## Negative log-likelihood in `TMB`

The Poisson HMM negative log-likelihood function in R can be written the following way.
```{Rcpp 3poi_hmm.cpp, code=readLines("code/poi_hmm.cpp"), eval=FALSE}
```

The case where $m = 1$ doesn't involve a hidden state, and thus is a Poisson regression instead of a Poisson HMM.
Nonetheless, `TMB` also accelerates its estimation and may be useful to the reader.
