--- 
title: "A gentle tutorial of accelerated parameter and confidence interval estimation for Hidden Markov Models using Template Model Builder"
author:
  - Timothée Bacri [timothee.bacri@uib.no](mailto:timothee.bacri@uib.no?subject=[GitHub%20TMB%20HMM])
  - Jan Bulla [jan.bulla@uib.no](mailto:jan.bulla@uib.no?subject=[GitHub%20TMB%20HMM])
  - Geir Berentsen [geir.berentsen@nhh.no](mailto:geir.berentsen@nhh.no?subject=[GitHub%20TMB%20HMM])
  - Sondre Hølleland [sondre.hoelleland@hi.no](mailto:sondre.hoelleland@hi.no?subject=[GitHub%20TMB%20HMM])
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [packages.bib, paper.bib]
biblio-style: apalike
# csl: apa.csl
link-citations: yes
description: "This accompanies the article published at [GITHUB LINK]."
url: "https://github.com/timothee-bacri/HMM_with_TMB"
github-repo: "timothee-bacri/HMM_with_TMB"
output:
  bookdown::gitbook:
    toc_depth: 4
    css: width.css
    split_bib: no
header-includes:
  - \usepackage{amsthm,amsmath,amssymb,amsfonts,amssymb}
  - \usepackage{listings}
  - \newcommand{\bs}[1]{{\boldsymbol#1}}
  - \DeclareMathOperator*{\argmax}{argmax}
editor_options:
  markdown:
    wrap: 80
---

# Introduction
<!-- COMPILE WITH bookdown::render_book("index.Rmd") -->
Welcome !

This website aims to accompany the reader of [Link to official article]

The files used in this repository are available in *https://github.com/timothee-bacri/HMM_with_TMB*.

For a description of the files and the directory structure, please read [Directory Structure].\
Note that only the folders [code/](#code), [functions/](#functions), and [data/](#data) contain files used in the article.
The other folders and files relate to this Gitbook and are not described in [Directory Structure].

For a description of the functions in *[functions/utils.R](#utils.R)* and their parameters and return objects, please take a look at *[functions/utils.R-explanation.R](#utils.R-explanation.R)*.

```{r 0import-files, cache = FALSE, include = FALSE}
library(knitr)
source('code/setup_parameters.R')
knitr::read_chunk('functions/utils.R')
knitr::read_chunk('functions/norm_utils.R')
knitr::read_chunk('functions/mvnorm_utils.R')
knitr::opts_chunk$set(collapse = TRUE)
```

```{r 0packages_in_bib, include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(packages, 'packages.bib')
```

<!-- # License {-} -->
<!-- This project is licensed under the xxxx License - see [LICENSE.md](LICENSE.md) for details. -->
