--- 
title: "A gentle tutorial of accelerated parameter and confidence interval estimation for Hidden Markov Models using Template Model Builder"
author:
  - Timothée Bacri [timothee.bacri@uib.no](mailto:timothee.bacri@uib.no?subject=[GitHub%20TMB%20HMM])
  - Jan Bulla [jan.bulla@uib.no](mailto:jan.bulla@uib.no?subject=[GitHub%20TMB%20HMM])
  - Geir Berentsen [geir.berentsen@nhh.no](mailto:geir.berentsen@nhh.no?subject=[GitHub%20TMB%20HMM])
  - Sondre Hølleland [sondre.hoelleland@hi.no](mailto:sondre.hoelleland@hi.no?subject=[GitHub%20TMB%20HMM])
# date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [packages.bib, paper.bib]
biblio-style: apalike
# csl: apa.csl
link-citations: yes
description: "This accompanies the article published at [GITHUB LINK]."
url: "https://github.com/timothee-bacri/HMM_with_TMB"
github-repo: "timothee-bacri/HMM_with_TMB"
output:
  bookdown::gitbook:
    # https://github.com/bblodfon/rtemps
    includes:
      in_header: hide_code.html # inspired by: https://stackoverflow.com/questions/45360998/code-folding-in-bookdown
    toc_depth: 4
    css:
      - width.css
      - style.css
    split_bib: no
    config:
      download: no
      sharing:
        # github: yes
        # twitter: yes
        # facebook: yes
        all: yes
header-includes:
  - \usepackage{amsthm,amsmath,amssymb,amsfonts,amssymb}
  - \usepackage{listings}
  - \newcommand{\bs}[1]{{\boldsymbol#1}}
  - \DeclareMathOperator*{\argmax}{argmax}
editor_options:
  markdown:
    wrap: 80
---

# Introduction
<!-- COMPILE WITH bookdown::render_book("index.Rmd") -->
Welcome !

This website aims to accompany the reader of *https://doi.org/10.1002/bimj.202100256*.

The files used in this repository are available in *https://github.com/timothee-bacri/HMM_with_TMB*.

For a description of the files and the directory structure, please read [Directory Structure].\
Note that only the folders [code/](#code), [functions/](#functions), and [data/](#data) contain files used in the article.
The other folders and files relate to this Gitbook and are not described in [Directory Structure].

For a description of the functions in *[functions/utils.R](#utils.R)* and their parameters and return objects, please take a look at *[functions/utils.R-explanation.R](#utils.R-explanation.R)*.

Additionally, as a by-product of using `TMB`, we can derive standard errors and hence confidence intervals for smoothing probabilities (as defined in [Section 11](#local-decoding-cis)).

## Other article

A preprint version of *https://doi.org/10.1080/00949655.2023.2226788* that is published can be found along with the code of its results in the folder [paper2-github-preprint/](https://github.com/timothee-bacri/HMM_with_TMB/blob/main/paper2-github-preprint).

This article looks at HMM optimization using popular optimization functions, compares their performance along with their robustness to bad initial parameters, and presents and analyzes smoothing probabilities confidence intervals.
Its code is organized in a similar fashion to the article that this repository accompanies.

```{r 0import-files, cache = FALSE, include = FALSE}
library(knitr)
source('code/setup_parameters.R')
knitr::read_chunk('functions/utils.R')
knitr::read_chunk('functions/norm_utils.R')
knitr::read_chunk('functions/mvnorm_utils.R')
knitr::opts_chunk$set(collapse = TRUE)
```

```{r 0packages_in_bib, include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(packages, 'packages.bib')
```

<!-- # License {-} -->
<!-- This project is licensed under the xxxx License - see [LICENSE.md](LICENSE.md) for details. -->
