--- 
title: "[ UNDER CONSTRUCTION ] A gentle tutorial of accelerated parameter and confidence interval estimation for Hidden Markov Models using Template Model Builder"
author:
  - Timothée Bacri [timothee.bacri@uib.no](mailto:timothee.bacri@uib.no?subject=[GitHub%20TMB%20HMM])
  - Jan Bulla [jan.bulla@uib.no](mailto:jan.bulla@uib.no?subject=[GitHub%20TMB%20HMM])
  - Geir Berentsen [geir.berentsen@nhh.no](mailto:geir.berentsen@nhh.no?subject=[GitHub%20TMB%20HMM])
  - Sondre Hølleland [sondre.hoelleland@hi.no](mailto:sondre.hoelleland@hi.no?subject=[GitHub%20TMB%20HMM])
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib, paper.bib]
biblio-style: apalike
link-citations: yes
description: "This accompanies the article published at [GITHUB LINK]."
url: "https://github.com/timothee-bacri/HMM_with_TMB"
github-repo: "timothee-bacri/HMM_with_TMB"
output:
  # html_document:
  #   toc: true
  #   toc_float:
  #     collapsed: false
  #     smooth_scroll: true
  # https://bookdown.org/yihui/bookdown/html.html#gitbook-style
  bookdown::gitbook:
    collapse: none
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
header-includes:
  - \usepackage{amsthm,amsmath,amssymb,amsfonts,amssymb}
  - \usepackage{listings}
  - \newcommand{\b}[1]{{\boldsymbol#1}}
  # - \newcommand{\balpha}{{\boldsymbol \alpha}}
  # - \newcommand{\bphi}{{\boldsymbol \varphi}}
  # - \newcommand{\bone}{{\boldsymbol 1}}
  # - \newcommand{\bgamma}{{\boldsymbol \gamma}}
  # - \newcommand{\bp}{{\boldsymbol P}}
  # - \newcommand{\bx}{{\boldsymbol x}}
  # - \newcommand{\by}{{\boldsymbol y}}
  - \DeclareMathOperator*{\argmax}{argmax}
---

# Summary {-}
<!-- bookdown::render_book("index.Rmd") -->
Welcome !

This website aims to accompany the reader of [Link to official article]

The files used in this repository are available in *https://github.com/timothee-bacri/HMM_with_TMB*.

For a description of the files and the directory structure, please read [Directory Structure].\
Note that only the folders [code](#code/), [functions](#functions), and [data](#data) contain files used in the article.
The other folders and files relate to this Gitbook and are not described in [Directory Structure].

```{r 0import-files, echo = FALSE, cache = FALSE, message = FALSE, results = FALSE}
library(knitr)
suppressMessages(source("code/packages.R"))
knitr::read_chunk('functions/utils.R')
source('code/setup_parameters.R')
```

```{r 0packages_in_bib, include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

# License
This project is licensed under the xxxx License - see [LICENSE.md](LICENSE.md) for details.
